---
title: "lab7withwork"
author: "Rachel Behm"
date: "February 21, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



Packages
```{r}
library(tidyverse)
library(tmap)
library(sf)
library(spatstat)
library(maptools)
library(sp)
library(raster)
library(gstat)
```

#1. Hawaii Raster data
```{r}
#bring in the tiffs (sf dont work for raster data gotta use raster package)

hi_par <-raster("PAR_CLIM_M.tif")

plot(hi_par) #wow~ its hawaii shaped and has info


hi_sst <-raster("SST_LTM.tif")
plot(hi_sst)


hi_chl <-raster("CHL_LTM.tif")
plot(hi_chl)

#if want to look at them side by side...
par(mfrow = c(1,3))
plot(hi_par)
plot(hi_sst)
plot(hi_chl)

#wowee zowie

#if call name of raster in console, shows raster info like cells and resolution etc
#gotta update datum since not wgs84
#cann use @ symbol to look at each individual part ex) hi_sst@crs
```

Reproject to WGS84
```{r}
#gotta find the secret R code (down below) for wgs84 via google----SPACING MATTERS!!!!!!!!!

wgs84<- "+proj=longlat +datum=WGS84 +ellips=WGS84 +no_defs"
#so we dont have to keep retyping it


#now reproject via projectRaster, method = bilinear since continuous data
hi_sst_84 <- projectRaster(hi_sst, crs = wgs84, method = "bilinear")
#if we check via @ we can see the datum has changed and the extent looks like normal latlong

plot(hi_sst_84)
#looks the same lol

#but what if we wanna cut it and make it less res or smaller or show a smaller scope
#decrease resolution by resample via aggregate (cells)


```

use aggregate (raster package) for resampling
```{r}
#example
sst_rs<- aggregate(hi_sst, fact=10)
plot(sst_rs)
#resolution decrease by order of 10
#really only needed if super massive(instead of running the biiiiig thing)
```

crop it like its hot
```{r}
#cropping rasters (changing the extents)
hi_sst_84@extent
#all of hawaii, big box

#create box first then crop
#these extent values come from allison, you would find what u need yourself
bounds <- as(extent(-156.2, -154.5, 18.7, 20.4), "SpatialPolygons")
#use these numbers to make an extent to save AS a spatial polygon named

#theres no crs for the polygon yet so we have to set it now- have to do before crop
#crs works if no crs but if already has then use project raster
crs(bounds) <- crs(hi_sst_84)

#now we can see has wgs84

#############################################################################################

#Now crop
sst_crop<- crop(hi_sst_84,bounds)
plot(sst_crop)
#now we see just the one island
```

algebra with rasters
```{r}
#lets make up a fake variable we want to map
#fake variable = tropicality which is sum of par+sst+chlA*2

hi_par_84<- projectRaster(hi_par,crs =wgs84, method ="bilinear")

#########################################################################################
#note: for thematic or categorical raster data use method = "ngm" (nearest neighboor) instead of bilinear 
#########################################################################################

hi_chla_84<- projectRaster(hi_chl,crs =wgs84, method ="bilinear")

par(mfrow= c(1,3))
plot(hi_sst_84)
plot(hi_par_84)
plot(hi_chla_84)

#notice how different the scales are - give insight into what tropicalities scale would be like

#########################################################################################
trop <- hi_par_84 +hi_sst_84 + 2*hi_chla_84
#plot intersection- not perfectly additive, diff col and rows

plot(trop)
```

make nicer plots with tmap
```{r}
#to call stuff out of folders... 
islands <- read_sf(dsn = 'islands', layer = "Island_boundaries") %>%
  dplyr::select(Island) %>%
  st_simplify(dTolerance=10) %>%
  st_transform(crs = 4326)
#dsn = new directory, layer = use this one
#have to specify which package want to use select from since do diff stuff
plot(islands)
#one island is missing lol

############################################################################################
#now lets look at it in tmap 

#for static viewing
tmap_mode("plot")
#for interactive viewing
#tmap_mode("plot")

sst_map <- tm_shape(hi_sst_84) +
  tm_raster(title = "Mean Sea Surface Temperature") + 
  tm_layout(bg.color ="navyblue",
            legend.position = c("left", "bottom"),
            legend.text.color = "white",
            legend.text.size = 0.5) +
#now we gonna add another shape of islands data
  tm_shape(islands) +
  tm_fill ("darkgreen")

#can see far left one is empty thats because that island was missing in the vector data we had for islands
########################################################################################

#to save what we made
tmap_save(sst_map, "allison_hawaii.png", height = 5)
#height is in inches, how big to make pic
#now this file is in our project folder
```

conditional rasters and masking
```{r}
#extents that aree the same and same col&rows IMPORTANT

#ex) looking for potential habitat of rare species with sst preferred range of >= 25.6 degrees and PAR below 54


#if we look at PAR and SST we can see the col#'s are diff and extents slightly different
#####################################################################################
# to change extent

extent(hi_sst_84) <- extent(hi_par_84) #easy peexy

# can double check via...
compareRaster(hi_sst_84, hi_par_84)
#tells us columns still different


#the info below is copied from the info for the hi_sst_84 raster
cr<- raster(nrow= 822,
            ncol =1229,
            xmn = -160.4365,
            xmx= -154.5373,
            ymn  =18.7309,
            ymx = 22.44634)

sst_new <- resample(hi_sst_84, cr, method= "bilinear")
compareRaster(sst_new, hi_par_84)

# yay, they match

```

cropped version just for kauai
```{r}
bounds_main <- as(extent(-159.9,-159.2, 21.7,22.3), 'SpatialPolygons')

crs(bounds_main) <- crs(sst_new)

#now crop with new bounds

par_kauai <- crop(hi_par_84, bounds_main)
sst_kauai <- crop(sst_new, bounds_main)

plot(par_kauai)
plot(sst_kauai)
############################################################################################
```

isolate regions of particular conditions
```{r}
#want temp >= 25.3 and par < 53
par_hab <- par_kauai #safety copy

#make these plots disappear
par_hab[par_hab >= 54] <- NA

plot(par_hab)


sst_hab<- sst_kauai
sst_hab[sst_hab < 25.4] <- NA
########################################################################################
#where do these verlap
suit_hab <- mask(sst_hab, par_hab)
plot(suit_hab)
```

#2. Point pattern analysis
red tree voles in humboldt county

```{r}

voles <- read_sf(dsn= 'redtreevoledata', layer = "ds033") %>%
  dplyr::select(COUNTY) %>%
  filter(COUNTY == "HUM") %>%
  st_transform(crs = 4326)


#vole observations
plot(voles)

#now lets get the county outline
#NOTE WILL WANT TO USE THIS SHAPEFILE FOR THE HW
humboldt <- read_sf(dsn = 'redtreevoledata', layer ="california_county_shape_file") %>%
  filter(NAME == "Humboldt") %>%
  dplyr::select(NAME) 

plot(humboldt)

#add a projection
st_crs(humboldt)<- 4326


#plot together via tmap
tm_shape(humboldt) + 
  tm_fill() +
  tm_shape(voles) + 
  tm_dots(size = 0.2)

#now are they clustered or random or what?

#can also plot with ggplot
ggplot()+
  geom_sf(data=humboldt) +
  geom_sf(data= voles)

#can save with ggsave
ggsave("humvoles.png", 
       units="in",
       width = 4, 
       height = 6, 
       dpi = 300)
```

lets go exploringgggg(point pattern analysis)
quadrat analysis vs distance based (neighboor, G&K) 
```{r}
#must convert from sf to regular obj since sf doesnt work
voles_sp <- as(voles, "Spatial")
voles_ppp <- as(voles_sp, "ppp")
#ppp point pattern object

humboldt_sp <-as(humboldt,"Spatial")
humboldt_win <- as(humboldt_sp, "owin")
#turn humboldt outline into the pp window

###########################################################################################
#combine together points and window (via lat lon (stored as x and y)) 
voles_pb <- ppp(voles_ppp$x, voles_ppp$y, window = humboldt_win)

plot(voles_pb)
#wowie stored together and plotted together woot
```

quadrat test
```{r}
#testing for spatial evenness (not CSR)

#give col and rows to split the map for qt
vole_qt<- quadrat.test(voles_pb, nx= 5, ny=10)
#warning tells you that breaking it up might give small  counts in those quadrats

vole_qt
#p is very small, reject null of spatial evenness therefore NOT EVEN

plot(voles_pb)
plot(vole_qt, add = TRUE, cex = 0.4)
#ooo very cool shows the polygons
```

plot kernel densities for spatial data:
```{r}
point_density <- density(voles_pb, sigma = 0.02)
plot(point_density)
#density is risky because bandwidth make data diff- can mislead

######################################################################################
#now lets make that as a raster and look at it
vole_raster <- raster(point_density, crs = wgs84)

tm_shape(vole_raster) + 
  tm_raster()
  
```

nearest neighboor 
G- dist of each to its NEAREST neighbor
K- how close all neighboring obs are to an event (concentric circles)


G -FUCNCTION
```{r}
#create sequence of distances to calculate proportions
#these numbers from allison and from testing
r <- seq(0,0.15, by = 0.005)
#at top everyone should be in that one

#assume same number of events to compare to csr
#envelope our data
#Gest = gestimate (what we wanna use)
#r = distances we set to r
#nsim = number of simulations
gfunction <- envelope(voles_pb, fun = Gest, r = r, nsim = 20) 

plot(gfunction)
plot(gfunction$obs ~ gfunction$r, type ="l", col ="black")
#type = l means line
#this is the same thing we did in lecture on tuesday but did it all for us 

#add in theoretical (CSR) too so can compare
plot(gfunction$obs ~ gfunction$r, type ="l", col ="black")
lines(gfunction$theo ~ gfunction$r, type ="l", col ="red")

#why do they look different?
#obs gets steeper quicker = MORE CLUSTERED THAN THEORETICAL(CSR)

```


K/L - function (theyre the same) -how close are ALL neighboors to all events in window
```{r}
#sequence of distances
#greater than in g function since need to look at bigger space since want all
r2 <- seq(0,0.5, by = 0.05)

#make new function same way
#Lest= l estimate 
#global= true means apply over entire window
lfunction <- envelope(voles_pb, fun = Lest, r = r2, nsim = 20, global = TRUE)

#now we plot like before (observed vs csr)
plot(lfunction$obs ~ lfunction$r, type = "l", col = "blue")
lines(lfunction$theo ~ lfunction$r, type = "l", col = "red")


#at lower dist, greater intensity surrounding event of ours compared to csr = CLUSTERING

```


Hypothesis test for CSR: Diggle-Cressie-Loosmore-Ford test of CSR
```{r}

DCLFTest <- dclf.test(voles_pb, nsim = 30) 
DCLFTest

#can see p value is significant and statistically not CSR 

#remeber p value can be skewed by sampling size

```



#Part 3.Spatial interpolation with krigin with Kansas rainfall 
```{r}
ks_rain <- read_csv("KSRain2.csv")
#now we gotta convert their lat and lon to spatial info in R cause it doesnt know

#remember make sure lon goes first, can add in crs here too woo
ks_sf  <-  st_as_sf(ks_rain, coords = c("LON", "LAT"), 
                 crs = 4326)

plot(ks_sf)

#we are just interested in rainful
```
```{r}
#get county info
# gotta change dsn cause in a folder
ks_counties <- read_sf(dsn = 'KSCounties', layer = "ks_counties_shapefile")

#has coords cause shapefile now just gotta add projection
st_crs(ks_counties) = 4326

plot(ks_counties)

#now lets make a tmap plot, can call attribute in the dots (rainfall =tmp)
tm_shape(ks_counties) +
  tm_fill()+
  tm_shape(ks_sf)+
  tm_dots("AMT", size = 0.5)

#will probably have highest error in places with lowest number of predictions (can see blank spots)
```


But we want to make predictions across the entire state using kriging. 
First, make the rainfall data a Spatial Points data frame: 
```{r}
ks_sp  <- as_Spatial(ks_sf)
#now the class has been changed (theres two ways to change to spatial this is the other way)

#now we gotta create a grid to find predicted values and variogram to find weighting
```

Make spatial grid to interpolate values over
```{r}
#allison chose these resolution wise for what makes sense for the bounds
lat <- seq(37, 40, length.out = 200)
long <- seq(-94.6,-102, length.out = 200)

#take these two vectors and turn it into a grid
grid <- expand.grid(lon = long, lat = lat)
#set lat and lon as ours

grid_sf <- st_as_sf(grid, coords = c("lon","lat"), crs = 4326)
#now convert to sf object with coords and as wgs84= MAKES SPATIAL


grid_sp <- as_Spatial(grid_sf)
#spatial stats package dumb so have to convert to their spatial object so can do stats


#ok now have the grid now find weighting

```

make variogram and find variogram model
```{r}

# Create the variogram:
ks_vgm <- variogram(AMT ~ 1, ks_sp)
#as a function of 1 means ordinary kriging cause we dont know stationary mean
plot(ks_vgm)
#if totally independent cant make predictions, we can see that there is a correlation
#this var shows how that changes over distance and how weighting decays

#now lets turn this into a model so can do functions


ks_vgm_fit <- fit.variogram(ks_vgm, model = vgm(nugget = 0.2, psill = 0.8, range = 200, model = "Sph"))
#model = vgm = make variogram
#get #'s via 
#range= stop changing
#just need starting estimate
#model = sph = spherical  - diff fits we just using this one right now (allison told us to use this one)


# Plot them both together
plot(ks_vgm, ks_vgm_fit) # Cool! So what are the values
#solid line = fitted, looks pretty good
```

Now, kriging! (interpolation time)
```{r}

ks_krige <- krige(AMT ~ 1, ks_sp, grid_sp, model=ks_vgm_fit)
#function of 1 means ordinary kriging
#krige(attribute ~ type of krige, data, grid, variogram)


#yucky to view so lets make df so can view
ks_krige_df <- as.data.frame(ks_krige) 
#then view in console

#rename  columns cause names bad
ks_krige_2 <- ks_krige_df %>% 
  rename(lon = coords.x1, lat = coords.x2, predicted = var1.pred, err = var1.var)
#view in console


#can convert to spatial data since it has lat lon (we renamed to that)
rain_predicted  <-  st_as_sf(ks_krige_2, coords = c("lon", "lat"), 
                 crs = 4326)

#################################################################################################
# Get Kansas outline (so can crop to it)
ks <- read_sf(dsn = "states", layer = "cb_2017_us_state_20m") %>% 
  dplyr::select(NAME) %>% 
  filter(NAME == "Kansas") %>% 
  st_transform(crs = 4326)

plot(rain_predicted)
#shows over whole grid which is just rectangle now want to crop so just show kansas

# Crop the rainfall data
rain_cropped <- st_intersection(rain_predicted, ks)

plot(rain_cropped) 
#can see that the corner is gone so its actually kansas

```

#hw3 :
1.
read papers and write about em
2. timeseries
3. map national parks
have to learn how to add text labels to a map she wont tell you how to
4. lizards - advice is dont make it more difficult than it is - MAKE IT SIMPLE
 
 due in 2 tuesdays from now